---
title: "Introduction to R Programming and R Markdown"
author: "SMUREU Data Science Bootcamp"
output: html_notebook
---

# Introduction to R Programming

This notebook is the R companion to the Python introduction.

The goal is not to memorize every command.

The goal is to become comfortable reading, running, modifying, and experimenting with code.

By the end of this notebook you should be able to:

-   run R code in an R Markdown notebook
-   create variables
-   work with vectors and lists
-   use data frames and tibbles
-   filter, sort, summarize, and join data
-   make a simple plot

```{r setup}
library(tidyverse)
```

# Operating This Notebook

R Markdown notebooks combine:

-   text
-   code
-   output
-   plots
-   notes

Code appears in chunks like this:

```{r}
1 + 1
```

You can run a chunk using the run button or the keyboard shortcut in your editor.

If something breaks:

1.  Save your work
2.  Restart the R session
3.  Run the chunks again from the top

# Basic Math

R can do calculations like a calculator.

```{r}
1 + 1
```

```{r}
2.5 - 2.0
```

```{r}
10 * 5
```

```{r}
100 / 4
```

# Variables

Variables allow us to save values for later.

```{r}
dallas_population <- 1300000
smu_students <- 12000
```

```{r}
dallas_population + smu_students
```

R commonly uses `<-` for assignment.

You will also see `=` used, but `<-` is the traditional R style.

# Strings

Strings allow us to store text.

```{r}
our_string <- "Here is a string in R!"
our_string
```

```{r}
print(our_string)
```

# Vectors

A vector is a collection of values of the same general type.

```{r}
favorite_foods <- c("tacos", "ramen", "pizza")
favorite_foods
```

R indexing starts at 1.

```{r}
favorite_foods[1]
```

```{r}
favorite_foods[length(favorite_foods)]
```

## You Try

Create a vector containing:

-   your hometown
-   your favorite movie
-   your favorite food

Then print the first and last item.

```{r}
# Create your vector here

# Print the first item

# Print the last item
```

# Named Vectors

Named vectors let us attach labels to values.

```{r}
city_populations <- c(
  Dallas = 1345000,
  Taos = 5668,
  Houston = 2326000
)

city_populations
```

```{r}
city_populations["Taos"]
```

# Libraries

R libraries are collections of pre-written tools.

Instead of writing everything from scratch, we can load libraries that help us:

-   analyze data
-   make plots
-   work with maps
-   clean text

One of the most important collections of libraries for data science in R is the `tidyverse`.

```{r}
library(tidyverse)
```

# Data Frames and Tibbles

A data frame is a table.

It has rows and columns, like a spreadsheet.

A tibble is a modern tidyverse-friendly version of a data frame.

```{r}
books <- tibble(
  Books = c(
    "The Silmarillion",
    "The Hobbit",
    "The Fellowship of the Ring",
    "The Two Towers",
    "The Return of the King"
  ),
  Words = c(130115, 95506, 187726, 156147, 137037)
)

books
```

# Selecting Columns

```{r}
books$Books
```

```{r}
books |> select(Books)
```

# Selecting Rows

```{r}
books |> slice(1:2)
```

```{r}
books |> slice(3:4)
```

# Selecting Rows and Columns

```{r}
books |> 
  slice(3:4) |> 
  select(Words)
```

# Operating on Columns

```{r}
total_words <- sum(books$Words)
total_words
```

## You Try

1.  Select only the `Words` column
2.  Select the first 3 rows
3.  Find the total number of words across all books

```{r}
# Your code here
```

# Loops and Functions

Computers are good at doing repeated tasks.

Here is a simple loop.

```{r}
for (book in books$Books) {
  print(book)
}
```

Functions let us save a reusable set of instructions.

```{r}
square_me <- function(n) {
  n * n
}
```

```{r}
square_me(4)
```

# More Complex Data

Below we have chapter-level word counts from The Hobbit and The Lord of the Rings.

```{r}
chapters <- tribble(
  ~CollectionNum, ~ChapterNum, ~ChapterName, ~WordCount, ~BookNum,
  0, 1, "An Unexpected Party", 8638, 0,
  0, 2, "Roast Mutton", 5257, 0,
  0, 3, "A Short Rest", 2876, 0,
  0, 4, "Over Hill and Under Hill", 4034, 0,
  0, 5, "Riddles in the Dark", 6967, 0,
  0, 6, "Out of the Frying Pan into the Fire", 6703, 0,
  0, 7, "Queer Lodgings", 9027, 0,
  0, 8, "Flies and Spiders", 10223, 0,
  0, 9, "Barrels Out of Bond", 5833, 0,
  0, 10, "A Warm Welcome", 3930, 0,
  0, 11, "On the Doorstep", 3001, 0,
  0, 12, "Inside Information", 7132, 0,
  0, 13, "Not At Home", 3909, 0,
  0, 14, "Fire and Water", 3236, 0,
  0, 15, "The Gathering of the Clouds", 3362, 0,
  0, 16, "A Thief in the Night", 2153, 0,
  0, 17, "The Clouds Burst", 3949, 0,
  0, 18, "The Return Journey", 2815, 0,
  0, 19, "The Last Stage", 2461, 0,
  1, -4, "Concerning Hobbits", 3406, 1,
  1, -3, "Concerning Pipeweed", 600, 1,
  1, -2, "Of the Ordering of the Shire", 2431, 1,
  1, -1, "Note on the Shire Records", 914, 1,
  1, 1, "A Long-expected Party", 10012, 1,
  1, 2, "The Shadow of the Past", 11311, 1,
  1, 3, "Three is Company", 9763, 1,
  1, 4, "A Short Cut to Mushrooms", 5957, 1,
  1, 5, "A Conspiracy Unmasked", 5196, 1,
  1, 6, "The Old Forest", 6502, 1,
  1, 7, "In the House of Tom Bombadil", 5501, 1,
  1, 8, "Fog on the Barrow-downs", 6694, 1,
  1, 9, "At the Sign of the Prancing Pony", 6251, 1,
  1, 10, "Strider", 5905, 1,
  1, 11, "A Knife in the Dark", 9468, 1,
  1, 12, "Flight to the Ford", 8805, 1,
  1, 1, "Many Meetings", 9085, 2,
  1, 2, "The Council of Elrond", 16360, 2,
  1, 3, "The Ring goes South", 10656, 2,
  1, 4, "A Journey in the Dark", 11501, 2,
  1, 5, "The Bridge of Khazad-dum", 5428, 2,
  1, 6, "Lothlorien", 9387, 2,
  1, 7, "The Mirror of Galadriel", 6896, 2,
  1, 8, "Farewell to Lorien", 6174, 2,
  1, 9, "The Great River", 7218, 2,
  1, 10, "The Breaking of the Fellowship", 6305, 2,
  2, 1, "The Departure of Boromir", 3397, 3,
  2, 2, "The Riders of Rohan", 11133, 3,
  2, 3, "The Uruk-hai", 7854, 3,
  2, 4, "Treebeard", 12876, 3,
  2, 5, "The White Rider", 8856, 3,
  2, 6, "The King of the Golden Hall", 9303, 3,
  2, 7, "Helm's Deep", 7575, 3,
  2, 8, "The Road to Isengard", 7899, 3,
  2, 9, "Flotsam and Jetsam", 7789, 3,
  2, 10, "The Voice of Saruman", 5663, 3,
  2, 11, "The Palantir", 6325, 3,
  2, 1, "The Taming of Smeagol", 8375, 4,
  2, 2, "The Passage of the Marshes", 7357, 4,
  2, 3, "The Black Gate is Closed", 5881, 4,
  2, 4, "Of Herbs and Stewed Rabbit", 6975, 4,
  2, 5, "The Window on the West", 10120, 4,
  2, 6, "The Forbidden Pool", 5179, 4,
  2, 7, "Journey to the Crossroads", 4266, 4,
  2, 8, "The Stairs of Cirith Ungol", 6793, 4,
  2, 9, "Shelob's Lair", 5209, 4,
  2, 10, "The Choices of Master Samwise", 7322, 4,
  3, 1, "Minas Tirith", 13100, 5,
  3, 2, "The Passing of the Grey Company", 8586, 5,
  3, 3, "The Muster of Rohan", 6951, 5,
  3, 4, "The Siege of Gondor", 11793, 5,
  3, 5, "The Ride of the Rohirrim", 4358, 5,
  3, 6, "The Battle of the Pelennor Fields", 5225, 5,
  3, 7, "The Pyre of Denethor", 3736, 5,
  3, 8, "The Houses of Healing", 6731, 5,
  3, 9, "The Last Debate", 5416, 5,
  3, 10, "The Black Gate Opens", 5204, 5,
  3, 1, "The Tower of Cirith Ungol", 9721, 6,
  3, 2, "The Land of Shadow", 8446, 6,
  3, 3, "Mount Doom", 7777, 6,
  3, 4, "The Field of Cormallen", 4721, 6,
  3, 5, "The Steward and the King", 7639, 6,
  3, 6, "Many Partings", 7440, 6,
  3, 7, "Homeward Bound", 4106, 6,
  3, 8, "The Scouring of the Shire", 11296, 6,
  3, 9, "The Grey Havens", 4791, 6
)

chapters
```

# Mapping Values

We can use a small lookup table to translate collection numbers into names.

```{r}
title_map <- tibble(
  CollectionNum = c(0, 1, 2, 3),
  CollectionName = c(
    "The Hobbit",
    "The Fellowship of the Ring",
    "The Two Towers",
    "The Return of the King"
  )
)

chapters <- chapters |>
  left_join(title_map, by = "CollectionNum")

chapters
```

# Grouping and Summarizing

`group_by()` lets us create groups.

`summarise()` lets us calculate values for each group.

```{r}
chapters |>
  group_by(CollectionName) |>
  summarise(
    chapter_count = n(),
    total_words = sum(WordCount),
    average_chapter_length = mean(WordCount)
  )
```

## You Try

Use `group_by()` and `summarise()` to find the average chapter length for each collection.

```{r}
# Your code here
```

# Filtering

Filtering lets us select rows based on conditions.

```{r}
books |> 
  filter(Books == "The Hobbit")
```

```{r}
books |> 
  filter(Books != "The Hobbit")
```

```{r}
books |> 
  filter(str_detect(Books, "ing"))
```

```{r}
books |>
  filter(str_detect(Books, "ing"), Words > 150000)
```

# Joining Data

Merging or joining lets us combine datasets.

```{r}
dates <- tibble(
  Books = books$Books,
  Year = c(1977, 1934, 1954, 1954, 1954)
)

dates
```

```{r}
merged <- books |>
  left_join(dates, by = "Books")

merged
```

# Plotting

R's `ggplot2` package is part of the tidyverse and is one of the most widely used visualization tools in data science.

```{r}
ggplot(books, aes(x = Books, y = Words)) +
  geom_col() +
  labs(
    title = "Word Counts for Tolkien Books",
    x = "Book",
    y = "Words"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## You Try

Try changing the plot to use `coord_flip()`.

```{r}
# Your code here
```

# Final Mini Challenge

Using the Tolkien data frames from this notebook:

1.  Find the longest book
2.  Create a plot of word counts
3.  Filter books with more than 150,000 words
4.  Find the average chapter length for each collection

# AI Assistant Checkpoint

Try asking an AI assistant:

-   "Explain what this dplyr pipeline is doing."
-   "Why does group_by need summarise?"
-   "How would I make this plot easier to read?"
-   "What is the difference between a vector and a data frame in R?"

# Wrapping Up

We covered:

-   R Markdown basics
-   variables
-   vectors
-   data frames
-   filtering
-   grouping
-   joining
-   plotting
