---
title: "Mapping with sf in R"
author: "SMUREU Data Science Bootcamp"
output: html_notebook
---

# Mapping with sf in R

So far we have explored data using tables, filtering, sorting, and charts.

But many datasets also have a geographic component.

Maps can reveal patterns that are difficult to see in tables or plots.

Today we will:

-   load geographic data
-   calculate spatial features like area and centroids
-   create buffers
-   test spatial relationships
-   merge maps with ACS and parks data
-   create choropleth maps
-   interpret geographic patterns across Texas counties

```{r setup}
library(tidyverse)
library(sf)
```

# Load Texas County Boundaries

The `sf` package lets us work with spatial data in R.

An `sf` object is like a data frame with a special geometry column.

```{r}
counties <- st_read("data/tx_counties.geojson", quiet = TRUE)

counties |> glimpse()
```

```{r}
counties |>
  select(COUNTY, FIPS, SQUARE_MIL, geometry) |>
  head()
```

# First Map

```{r}
ggplot(counties) +
  geom_sf() +
  labs(title = "Texas Counties")
```

# Coordinate Systems and Measurement

The original county file uses latitude and longitude.

That is good for locating places on Earth, but not ideal for measuring area, distance, or buffers.

For measurement, we project the data into a coordinate system designed for Texas.

```{r}
counties_projected <- st_transform(counties, 3083)
```

# Area

`sf` calculates area using the units of the coordinate system.

After projecting to EPSG:3083, distances are measured in meters, so areas are measured in square meters.

We divide by 2,589,988.11 to convert square meters into square miles.

```{r}
SQ_METERS_PER_SQ_MILE <- 2589988.11

counties_projected <- counties_projected |>
  mutate(area_sq_miles = as.numeric(st_area(geometry)) / SQ_METERS_PER_SQ_MILE)
```

```{r}
counties_projected |>
  st_drop_geometry() |>
  select(COUNTY, area_sq_miles) |>
  arrange(desc(area_sq_miles)) |>
  head(10)
```

# Boundaries and Centroids

A centroid is the geometric center of a shape.

```{r}
county_centroids <- counties_projected |>
  mutate(geometry = st_centroid(geometry))
```

```{r}
ggplot() +
  geom_sf(data = counties_projected, fill = "white", color = "black") +
  geom_sf(data = county_centroids, color = "red", size = 0.5) +
  labs(title = "Texas County Centroids")
```

# Distance

Let's measure distance from Dallas County to every other county centroid.

```{r}
dallas_centroid <- county_centroids |>
  filter(COUNTY == "Dallas County") |>
  st_geometry()

county_centroids <- county_centroids |>
  mutate(distance_from_dallas_km = as.numeric(st_distance(geometry, dallas_centroid)) / 1000)
```

```{r}
county_centroids |>
  st_drop_geometry() |>
  select(COUNTY, distance_from_dallas_km) |>
  arrange(distance_from_dallas_km) |>
  head(10)
```

# Buffers

A buffer creates a zone around a geometry.

Here we create a 100 km buffer around Dallas County.

```{r}
dallas_county <- counties_projected |>
  filter(COUNTY == "Dallas County")

dallas_buffer_100km <- st_buffer(dallas_county, dist = 100000)
```

```{r}
ggplot() +
  geom_sf(data = counties_projected, fill = "white", color = "lightgray") +
  geom_sf(data = dallas_county, fill = "orange", color = "black") +
  geom_sf(data = dallas_buffer_100km, fill = NA, color = "red", linewidth = 1) +
  labs(title = "100 km Buffer Around Dallas County")
```

# Spatial Relationships

We can ask which counties intersect the 100 km buffer.

```{r}
counties_projected <- counties_projected |>
  mutate(intersects_dallas_buffer = lengths(st_intersects(geometry, dallas_buffer_100km)) > 0)

counties_projected |>
  st_drop_geometry() |>
  filter(intersects_dallas_buffer) |>
  select(COUNTY) |>
  arrange(COUNTY)
```

# Merge ACS Data

Now let's bring in the ACS dataset from the previous session.

```{r}
acs <- read_csv("data/texas_ACS_2024_sample.csv", show_col_types = FALSE) |>
  rename(
    county = name,
    population = `Total Population`,
    labor_force = `Population 16 years and over in labor force`,
    bachelors_degree = `Population 25 years and over with Bachelors Degree or higher`,
    poverty_population = `Population 16 and over with income in the past 12 months below poverty level`,
    internet_households = `Households with Internet Subscription`,
    renters = `Total Population in Renter Occupied Housing`,
    median_income = `Median Household Income in the Past 12 Months (In 2024 Inflation-adjusted Dollars)`,
    owners = `Total Population in Owner Occupied Housing`
  ) |>
  mutate(
    county_fips = str_sub(geoid, -5),
    poverty_rate = poverty_population / labor_force,
    bachelors_rate = bachelors_degree / population,
    internet_rate = internet_households / (owners + renters),
    renter_rate = renters / (owners + renters)
  )
```

```{r}
mapped <- counties_projected |>
  mutate(county_fips = str_pad(as.character(FIPS), width = 5, pad = "0")) |>
  left_join(acs, by = "county_fips")

mapped |> glimpse()
```

# Choropleth Maps

A choropleth map colors regions based on a data value.

```{r}
ggplot(mapped) +
  geom_sf(aes(fill = median_income), color = NA) +
  scale_fill_viridis_c(option = "plasma", na.value = "lightgrey") +
  labs(
    title = "Median Household Income by Texas County",
    fill = "Median Income"
  )
```

```{r}
ggplot(mapped) +
  geom_sf(aes(fill = poverty_rate), color = NA) +
  scale_fill_viridis_c(na.value = "lightgrey") +
  labs(
    title = "Poverty Rate by Texas County",
    fill = "Poverty Rate"
  )
```

# Missing Data and Parks

Missing data matters in maps.

Now let's merge in the state parks dataset.

```{r}
parks <- read_csv("data/texas_state_parks_by_county.csv", show_col_types = FALSE)

parks |> head()
```

```{r}
mapped <- mapped |>
  left_join(parks, by = c("geoid" = "geoid"))

mapped |>
  st_drop_geometry() |>
  select(COUNTY, state_park_count) |>
  head()
```

Here a missing park value will be treated as a modeling decision.

We will assume missing park records mean zero state parks in that county.

```{r}
ggplot(mapped) +
  geom_sf(aes(fill = state_park_count), color = "white") +
  scale_fill_viridis_c(na.value = "lightgrey") +
  labs(
    title = "State Park Counts by County with Missing Values",
    fill = "Parks"
  )
```

```{r}
mapped <- mapped |>
  mutate(state_park_count = replace_na(state_park_count, 0))
```

```{r}
ggplot(mapped) +
  geom_sf(aes(fill = state_park_count), color = "white") +
  scale_fill_viridis_c() +
  labs(
    title = "State Park Counts by County",
    fill = "Parks"
  )
```

# Capstone Example: Parks Within 100 km of a County

Now let's tie several spatial ideas together.

Suppose we want to ask:

> How many state parks could someone plausibly drive to within 100 km of a given county?

This example uses:

-   projections
-   centroids
-   buffers
-   spatial relationships
-   missing data
-   counting and summarizing

## Important Limitation

Our parks dataset counts parks by county, but it does not include the exact latitude and longitude of each park.

So this is an approximation.

We will estimate nearby parks by counting parks in counties whose centroids fall within 100 km of the selected county.

A better version of this analysis would use the actual point locations of each state park.

Large counties may have parks far from the county centroid.

```{r}
selected_county <- "Dallas County"

target_county <- counties_projected |>
  filter(COUNTY == selected_county)

buffer_100km <- st_buffer(target_county, dist = 100000)
```

```{r}
parks_clean <- parks |>
  mutate(
    county_fips = str_sub(as.character(geoid), -5)
  ) |>
  select(county_fips, state_park_count)
```

```{r}
counties_with_parks <- counties_projected |>
  mutate(county_fips = str_pad(as.character(FIPS), width = 5, pad = "0")) |>
  left_join(
    parks_clean,
    by = "county_fips"
  ) |>
  mutate(
    state_park_count = replace_na(state_park_count, 0),
    centroid_geometry = st_centroid(geometry)
  )
```

Because `sf` objects have one active geometry column, we make a separate centroid object for the within-buffer calculation.

```{r}
centroids_with_parks <- counties_with_parks |>
  st_set_geometry("centroid_geometry")
centroids_with_parks$within_100km <- lengths(
  st_within(
    st_geometry(centroids_with_parks),
    st_geometry(buffer_100km)
  )
) > 0
```

```{r}
nearby_parks <- centroids_with_parks |>
  st_drop_geometry() |>
  filter(within_100km) |>
  summarise(total_parks = sum(state_park_count)) |>
  pull(total_parks)
nearby_parks
```

```{r}
centroids_with_parks |>
  st_drop_geometry() |>
  filter(within_100km) |>
  select(COUNTY, state_park_count) |>
  arrange(desc(state_park_count))
```

```{r}
nearby_counties <- counties_projected |>
  filter(COUNTY %in% centroids_with_parks$COUNTY[centroids_with_parks$within_100km])

ggplot() +
  geom_sf(data = counties_projected, fill = "white", color = "lightgray") +
  geom_sf(data = nearby_counties, fill = "lightblue", color = "black", alpha = 0.5) +
  geom_sf(data = target_county, fill = "orange", color = "black") +
  geom_sf(data = buffer_100km, fill = NA, color = "red", linewidth = 1) +
  geom_sf(data = centroids_with_parks, aes(size = state_park_count), color = "lightgreen", alpha = 0.7) +
  labs(
    title = paste("Estimated state parks within 100 km of", selected_county, "County:", nearby_parks),
    size = "Park Count"
  )
```

## Discussion

What assumptions did we make?

-   We used county centroids instead of exact park locations.
-   We counted all parks in a county if that county's centroid was within the buffer.
-   We used a straight-line distance, not actual driving distance.
-   We did not consider roads, traffic, park size, or park quality.

This is still useful because it shows how spatial analysis can help us build an approximate answer when perfect data is not available.

# Mini Challenge

Create one map that surprised you.

Your map should:

-   use one variable from the ACS or parks dataset
-   include a legend
-   have a title

Then write 2-3 sentences answering:

1.  What pattern do you see?
2.  Why might that pattern exist?
3.  What is one reason this map might be misleading?

# AI Assistant Checkpoint

Try asking an AI assistant:

-   "What does this sf map show?"
-   "Why do I need to project geometries before measuring distance?"
-   "What does a spatial buffer mean?"
-   "What assumptions are built into this analysis?"
