---
title: "Working with and Visualizing Data in R"
author: "SMUREU Data Science Bootcamp"
output: html_notebook
---

# Working with and Visualizing Data in R

In this notebook we will use R to:

-   work with tabular data
-   create new columns
-   filter and sort data
-   build a simple scoring model
-   create visualizations
-   explore ACS-style county data
-   merge an additional parks dataset

```{r setup}
library(tidyverse)
```

# Part 1: Apartment Hunting

Let's start with a dataset that is easy to understand.

Suppose you are looking for an apartment near campus.

Different apartments have different:

-   rent
-   size
-   distance
-   ratings
-   pet policies

How could we use data to compare them?

```{r}
apts <- tibble(
  Apartment = c(
    "Mockingbird Flats",
    "Highland House",
    "Campus View",
    "The Village",
    "Oak Lawn Lofts",
    "Greenville Commons",
    "Mustang Station",
    "The Hidden Gem"
  ),
  Rent = c(950, 1450, 1100, 1700, 2100, 1250, 800, 1350),
  Bedrooms = c(1, 2, 1, 3, 2, 2, 1, 2),
  Bathrooms = c(1, 2, 1, 2, 2, 1, 1, 2),
  Sqft = c(550, 980, 620, 1400, 1150, 870, 450, 760),
  Distance = c(0.4, 1.8, 0.2, 3.5, 2.2, 1.1, 0.1, 0.9),
  Rating = c(3.8, 4.6, 3.2, 4.9, 4.7, 4.1, 2.9, 4.8),
  pets_allowed = c(
    "cat only",
    "dog and cat",
    "no",
    "all pets",
    "dog and cat",
    "cat only",
    "no",
    "all pets"
  )
)

apts
```

## Question 1: Which apartment is cheapest?

```{r}
apts |>
  arrange(Rent)
```

## Question 2: Which apartment gives the most space per dollar?

```{r}
apts <- apts |>
  mutate(sqft_per_dollar = Sqft / Rent)

apts |>
  arrange(desc(sqft_per_dollar))
```

Would everyone agree this is the best apartment?

What is missing from this comparison?

## Question 3: Which apartments fit a student's needs?

Suppose we want at least 2 bedrooms and rent below 1500.

```{r}
apts |>
  filter(Bedrooms >= 2, Rent < 1500)
```

# Building a Simple Scoring Model

Now let's build a rough score.

This is not the correct answer.

It is one possible model.

```{r}
apts <- apts |>
  mutate(
    space_score = (Bedrooms + Bathrooms) * Sqft,
    value_score = space_score / Rent
  )
```

Smaller distances should receive larger scores, so we use the inverse of distance.

```{r}
apts <- apts |>
  mutate(
    distance_score = 1 / Distance,
    final_score = value_score * distance_score * Rating
  )

apts |>
  select(Apartment, Rent, Sqft, Distance, Rating, pets_allowed, final_score) |>
  arrange(desc(final_score))
```

## Discussion

-   Do you trust this ranking?
-   What variables did we ignore?
-   Should pets matter?
-   Should distance matter more or less?
-   What would you change?

# Visualizing Apartment Data

## Rent by Apartment

```{r}
ggplot(apts, aes(x = reorder(Apartment, Rent), y = Rent)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Apartment Rent",
    x = "Apartment",
    y = "Rent"
  )
```

## Rent vs Rating

```{r}
ggplot(apts, aes(x = Rent, y = Rating, label = Apartment)) +
  geom_point() +
  labs(
    title = "Rent and Apartment Rating",
    x = "Monthly Rent",
    y = "Rating"
  )
```

## Score by Apartment

```{r}
ggplot(apts, aes(x = reorder(Apartment, final_score), y = final_score)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Apartment Score from Our Model",
    x = "Apartment",
    y = "Score"
  )
```

# Part 2: Working with ACS Data

Now let's move from a toy dataset to a real public-interest dataset.

```{r}
acs <- read_csv("data/texas_ACS_2024_sample.csv", show_col_types = FALSE)

acs |> glimpse()
```

# Rename Columns

Real datasets often have long column names.

Let's rename them for readability.

```{r}
acs <- acs |>
  rename(
    county = name,
    population = `Total Population`,
    labor_force = `Population 16 years and over in labor force`,
    bachelors_degree = `Population 25 years and over with Bachelors Degree or higher`,
    poverty_population = `Population 16 and over with income in the past 12 months below poverty level`,
    internet_households = `Households with Internet Subscription`,
    renters = `Total Population in Renter Occupied Housing`,
    median_income = `Median Household Income in the Past 12 Months (In 2024 Inflation-adjusted Dollars)`,
    owners = `Total Population in Owner Occupied Housing`
  ) |>
  mutate(
    county_name = str_remove(county, ", TX"),
    county_fips = str_sub(geoid, -5)
  )

acs |> head()
```

# Looking Beyond Totals

Raw totals often tell us where the largest counties are.

```{r}
acs |>
  arrange(desc(population)) |>
  select(county, population) |>
  head(10)
```

Rates often tell us something different.

The rates below are simplified educational examples. In real ACS work, pay close attention to denominators.

```{r}
acs <- acs |>
  mutate(
    poverty_rate = poverty_population / labor_force,
    bachelors_rate = bachelors_degree / population,
    internet_rate = internet_households / (owners + renters),
    renter_rate = renters / (owners + renters)
  )

acs |>
  select(county, poverty_rate, bachelors_rate, internet_rate, renter_rate) |>
  head()
```

# Visualizing ACS Data

## Income and Education

```{r}
ggplot(acs, aes(x = bachelors_rate, y = median_income)) +
  geom_point() +
  labs(
    title = "Education and Median Income by Texas County",
    x = "Bachelor's Degree Rate",
    y = "Median Household Income"
  )
```

## Poverty and Internet Access

```{r}
ggplot(acs, aes(x = poverty_rate, y = internet_rate)) +
  geom_point() +
  labs(
    title = "Poverty and Internet Access by Texas County",
    x = "Poverty Rate",
    y = "Internet Access Rate"
  )
```

# Finding Unusual Counties

These thresholds are chosen simply to help us find unusual counties.

```{r}
acs |>
  filter(median_income > 80000, bachelors_rate < 0.20) |>
  select(county, median_income, bachelors_rate, poverty_rate)
```

# Part 3: Merging State Parks Data

Now let's bring in a second dataset.

```{r}
parks <- read_csv("data/texas_state_parks_by_county.csv", show_col_types = FALSE)

parks |> head()
```

```{r}
acs_parks <- acs |>
  left_join(parks, by = c("geoid" = "geoid")) |>
  mutate(state_park_count = replace_na(state_park_count, 0))

acs_parks |>
  select(county, state_park_count, median_income, poverty_rate) |>
  arrange(desc(state_park_count)) |>
  head(10)
```

## Do counties with more parks have higher income?

```{r}
ggplot(acs_parks, aes(x = state_park_count, y = median_income)) +
  geom_point() +
  labs(
    title = "State Parks and Median Income",
    x = "State Park Count",
    y = "Median Household Income"
  )
```

If a relationship exists, does that mean parks cause income?

# Mini Challenge

Choose one of the following.

## Starter

Find the top 10 counties by one variable.

## Intermediate

Create a scatter plot comparing two variables.

## Challenge

Merge parks with ACS data and investigate whether parks seem related to another county-level variable.

## Open-Ended

Find one pattern that surprised you and explain what you think might be happening.

# AI Assistant Checkpoint

Try asking an AI assistant:

-   "What does this plot suggest?"
-   "What are three possible explanations for this pattern?"
-   "Can you find evidence in the data that supports or contradicts this explanation?"
-   "How could this visualization be improved?"
